Extreme heterogeneity of influenza virus transcriptional load in single cells

### 2017-09-14 paper/paper.tex
added version of Figure 5A for each sample individually

\begin{abstract}
Viral infection can cause large changes in a cell's transcriptome, but these changes have mostly been studied by bulk measurements on many cells.
We use single-cell mRNA sequencing to examine the consequences of influenza virus infection in individual cells.
We find extremely wide cell-to-cell variation in viral gene expression -- viral transcripts compose no more than a few percent of total mRNA in most infected cells, but a few cells derive over half their mRNA from virus.
Many infected cells fail to express at least one viral gene, and this gene absence explains some of the variation in viral transcriptional load.
Activation of innate immune pathways is rare, even in cells that fail to express influenza virus's major interferon antagonist.
We identify some cellular genes that co-vary in abundance with the amount of viral mRNA.
Overall, our results highlight the extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
\end{abstract}

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A). 
Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express. 
The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.

@@

\section{Discussion}
We have quantified the total transcriptome composition of single cells infected with influenza virus.
While we certainly observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.

@@

@@

However, despite extensive heterogeneity, there seems to be robustness to the infectious lifecycle.
For example, influenza, regardless of the viral burden, exhibits relatively constrained ratios between the viral transcripts.
This suggests that the processes that are driving heterogeneity of influenza mRNA abundance act equivalently throughout the viral genome.
Potential mechanisms include components such as binding and entry kinetics that act on a virion and thus on all segments packaged therein, limitations of transcription rate that are dependent upon the same pool of cellular resources, or potentially even feedback-dependent mechanisms controlling viral transcription.

>>>>

\begin{abstract}
Viral infection can cause large changes in a cell's transcriptome, but these changes have mostly been studied by bulk measurements on many cells.
We use single-cell mRNA sequencing to examine the consequences of influenza virus infection in individual cells.
We find extremely wide cell-to-cell variation in viral gene expression -- viral transcripts compose no more than a few percent of total mRNA in most infected cells, but a few cells derive over half their mRNA from virus.
Some infected cells fail to express at least one viral gene, and this gene absence partially explains the variation in viral transcriptional load.
Activation of innate immune pathways is rare, even in cells that fail to express influenza virus's major interferon antagonist.
We identify some cellular genes that co-vary in abundance with the amount of viral mRNA.
Overall, our results highlight the extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
\end{abstract}

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in Figure~\ref{fig:fluburdenbyflugene}.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low for meaningful statistical analyses.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express(Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}). 
The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.

@@

\section{Discussion}
We have quantified the total transcriptome composition of single cells infected with influenza virus.
While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.

@@

However, the absence of specific viral genes in some cells explains only a fraction of the variation in amount of viral mRNA, which still varies by over ...

@@
